---
title: Executive Constraints and Civil Conflict Onset
author: J.L.A. Krusell
date: "`r format(Sys.time(), '%Y-%m-%d')`"
bibliography: library.bib
csl: assets/american-journal-of-political-science.csl
link-citations: true
documentclass: scrreprt
fontsize: 11pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
output:
    pdf_document:
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate"]
    html_document:
        smart: true
        toc: true
abstract: Lorem ipsum
---

[//]: # Setup R environment for entire doc
```{r include = F, results = "hide"}
knitr::opts_chunk$set(echo = F, eval = F)

library(dplyr)
library(thesis.utils)
```

[//]: # Force TOC on separate page
\newpage

# Introduction {-}

# Theoretical Overview {-}

## Institutions and Civil Conflict

This paper takes as a starting premise a definition of institutions as
being "systems of established and embedded social rules that structure
social interactions" [@Hodgson2006]. More informally, we can consider
them to be the "rules of the game", encompassing for example codified
laws, norms, and shared understandings, all of which can constrain or
enable behaviour by influencing the perceptions and preferences of
actors through incentives and deterrents.[^institutions] Furthermore,
over time, the habits formed from these systems create expectations of
stable patterns of behaviour thereby shaping the interactions between
actors by reducing uncertainty [@North1990]. Although there is a wide
variety of institutions, the present focus will be on the type of
institutions which govern political behaviour.

[^institutions]: None of which is meant to imply that institutions
    stand external to actor perceptions and preferences as unchanging
    features. Rather, institutions are social constructs subject to
    changing normative values and able to be reinforced or undermined
    by behavioural feedback loops. This latter point is especially
    interesting within the context of civil conflict considering how
    the development of various institutions and conflict incidence can
    by jointly determined; however, it remains outside the scope of
    this paper.

This exposition is relevant to the study of civil conflict for both
motivation-based and opportunity-based explanatory accounts. To the
extent that the onset of civil conflict is a longer process featuring
waves of escalation and de-escalation rather than a single one-off
event [@Florea2012; @Sambanis2004b], institutions are a key influence
on the constitutive interactions between state and non-state
actors. For example, they can shift the opportunity costs of violence
altering the mobilization capacity of would be rebel
groups. Institutions can also determine the extent to which the state
is seen as accountable to and representative of the larger population,
thereby influencing the legitimacy of anti-government violence.  And,
their quality can shape the capacity of the state to suppress or
co-opt potential opposition. Although the preceding examples are far
from forming an exhaustive list, there are thus a number of different
ways in which institutions can enter into theoretical discussions
about civil conflict.

Unsurprisingly, insofar as the role of institutions has been examined
by previous literature the focus has largely been on the broader
notion of democracy, which at a minimum is a system regulating access
to political power through popular representation determined by free
and competitive elections [@Schumpeter1962]. Democracy serves as a
proxy for a number of normatively desirable institutions and has been
hypothesised to have a pacifying domestic effect according to two
predominant mechanisms. The first is centered on the concept of
relative deprivation, or inequalities in value expectations and value
capabilities, as a primary source of discontent that can motivate
collective violence [@Gurr1970]. [^gurr] Democracy is thought to
decrease the amount of political discontent by providing avenues for
free expression and adjudicating group conflicts through non-violent
institutional means --- for example, through the regular occurrence of
free and fair elections [@Hegre2014]. The second mechanism assumes
that violent conflict is costly and that there exists a bargained
solution preferable to fighting, assuming that actors are rational and
able to internalize the cost of violence [@Fearon1995]. One of the
reasons why a bargained solution may be impossible is the risk of
future shifts in the distribution of power. More specifically, a
future change in favor of relative strength of one of the parties can
provide an incentive to later renege. This influences the present
degree to which concessions can be offered and parties are willing to
credibly commit [@Powell2006]. Within the domestic sphere, democracy
offers a possible solution by formalising the regular contest of power
with full enfranchisement, thereby providing a *de jure* guarantee for
the balance of power [@Origins2006], and should, likewise, decrease
the likelihood of civil conflict.

[^gurr]: Gurr explains value expectations as being that which we think
    we are entitled to and value capabilities as that which we think
    we are capable of attaining. As Gurr then states, relative
    deprivation is the perceived discrepancy between the two. The link
    to collective violence, however, is conditional on the discontent
    being politicized, by which it is meant that the source of the
    discrepancy is identified and a normative and utilitarian
    justification is given for violent action, before finally being
    actualized.  Quantitative studies tend to focus on the first
    stage, the propensity for political regimes to generate
    discontent, given the difficultly of measuring the variables that
    may affect politicisation and actualisation.

Despite the theoretical basis linking democracy to reduced risks for
civil conflict, the empirical results tend to be mixed. @Hegre2001,
for example, find that democracies are at a lower risk for civil
conflict; however, no more so than fully authoritarian
states. Instead, anocratic regimes in between democracy and
authoritarianism are considered to be the most prone. The explanatory
logic is that both democratic and fully authoritarian states tend to
be consolidated regimes. In the case of the former, this means the
presence of institutions conducive to conflict resolution such that
the opportunity for rebellion may be present, given the presence of
civil liberties conducive to organizing, but there is a notable lack
in motivation for collective violence. For the latter, authoritarian
regimes may produce a significant amount of political discontent;
however, the opportunity is lacking due to the state's capacity and
willingness to violently repress dissent. Anocratic regimes are by
their very nature transitory, partly open and partly repressive,
encouraging political discontent against the state without the
capacity to either accommodate or suppress opposition.

In contrast, studies emphasizing the importance of political and
economic opportunities for rebellion in understanding when and where
conflicts break out, such as @Fearon2003 and @Collier2009, find
democracy to be insignificant. According to their theoretical
expositions, democracy should be associated with higher levels of
political rights, respect for ethnic and religious minorities, and
less repression, thus being a proxy for the latent level of political
discontent against the state. However, the argument follows that even
if the number of political grievances vary across regime types, they
are still ubiquitous enough to be a poor explanation for why conflict
breaks out in only a relatively small subset of countries. Instead,
the determining factors for conflict onset are the structural
conditions conducive to rebellion.[^opportunities]

[^opportunities]: For Fearon and Laitin these structural conditions
    are related to the features which make rebellion militarily
    viable.  Specifically, there's a higher risk of conflict onset
    where state capacity is weak, principally understood in terms of
    state revenue and its impact on policing and counterinsurgency
    operations; large impoverished populations; and insurgents having
    a terrain advantage. Collier *et al* start from the premise that
    war is expensive and the primary impediment to a rebellion is its
    financial feasibility relative to the strength of the state. On
    the basis of their results they go so far as provide support to
    the proposition, "where a rebellion is financially and militarily
    feasible it will occur." Both approaches, however, reach the same
    conclusion concerning the empirical evidence surrounding the role
    of democracy.

To provide a more nuanced picture, an ongoing strain of research has
been dedicated to disaggregating measures of democracy into either
finer regime type categories or specific measures of individual
political institutions. The theories offered tend to follow a general
formula whereby formal lack or poor quality of some institution(s),
measured either directly or proxied by regime type, undermines the
capacity of the state to prevent armed insurgency and/or incentivises
opposition violence by restricting the opportunities for political
participation and increasing political discontent. For example,
@Carey2007 examines the executive selection process of Sub-Saharan
Africa and argues that executive elections, even when they are of poor
quality, reduce the risk of large-scale violent dissent. @Fjelde2010a
breaks down authoritarian regime types in Africa attributing the
significant variation in conflict risk to the differing institutional
configurations affecting a regime's ability to coerce or co-opt the
opposition. The conclusion is that previous indeterminate results
regarding the link between political institutions and conflict risk
may have been hamstrung by aggregate measures which poorly reflect the
within-group variation of political regimes. In a larger sample
including democratic polities, @Goldstone2010 create a regime
classification scheme based on measures for openness of executive
recruitment and competitiveness of political participation, as well as
factionalism in order to distinguish between partial democracies. They
offer a theory of civil conflict occurring in countries where elite
divisions undermine the capacity of the state to defeat an insurgency
and claim that a relatively parsimonious model including their regime
classification offers better predictive accuracy than previous
analyses and provides evidence that political institutions are a
dominant driving factor for both political instability and civil
conflict. Finally, @Bartusevic2018 construct a regime typology focused
specifically on electoral features, which are argued form the
essential core of the concept of democracy. Where there are
constraints on electoral contestation, the authors contend that such
constraints restrict the opportunities for non-violent political
participation thereby legitimising anti-government violence and
self-selecting recruits into rebel groups willing to engage in armed
conflict. Evidence is presented which the authors claim supports not
just the significance of regime type, but a negative monotonic
relationship between democracy and the risk of civil conflict.

Studies utilising individual measures beyond regime classification
have been largely focused on the concept of governance. For instance,
@Taydas2010 examine institutional quality as a means to capturing a
regime's governing capacity theorised to be linked to conflict onset
both in terms of political discontent --- a regime's inability to
govern effectively and respond to public needs leading to grievances
and a loss in legitimacy --- and political opportunities for rebellion
--- poor institutional quality undermines the state's ability to
handle a nascent, violent uprising. Specifically, they assess
corruption, rule of law, and bureaucratic quality and present evidence
that all three are negatively related to civil conflict
onset. Similarly, @Hegre2015a argue that the presence of formal
representative institutions does not fully capture the extent to which
a regime is capable of realizing policy goals benefiting the
public. Poor quality of government can therefore undermine the
legitimacy of a regime when citizens identify the state as responsible
for relative deprivation. In their empirical analysis, the authors
separate informal institutions related to good governance from formal
democratic institutions and find both important for reducing the
incidence of civil conflict, or in this specific case the recurrence
of conflict.

Taken together, this recent literature on civil conflict onset points
to the ongoing importance of political institutions. There remains,
however, significant work to be done identifying the causal mechanisms
and effects of specific institutions, especially in terms of further
disaggregating the concept of democracy beyond ever finer
classifications of regime type [@ari2016]. Furthermore, to the extent
that previous research has mostly focused on the effects of
institutions on would-be rebel motivation and/or political opportunity
structures, there is also a gap in terms of empirical analyses on how
institutions shape the behaviour of the state in the lead up to a
civil conflict. Even in strictly economic explanations of civil
conflict, giving explanatory primacy either to greed-based motivation
or economic conditions conducive for rebellion, the state is a key
actor. How a state responds to dissent structures the opportunity
costs of violence as well as the salience of calls to mobilization by
rebel leaders. This latter point is especially relevant in linking
regime legitimacy to conflict since public perception is shaped not
simply by a regime's formal structure, but by its conduct in repeated
interactions with the public. To this end, it is imperative to examine
the types of institutions which enable or constrain actions by the
state which exacerbate the risk of conflict onset beyond state
capacity.

There are of course a large number of institutions whose presence and
quality can influence the behaviour of the state. Given the importance
of the executive as a central authority, however, especially in times
of political crisis, institutions which affect the executive's
decision making should be pertinent. Chief among these is the idea of
executive constraints, or the formal constraints on executive power by
other the branches of government. By analyzing executive constraints
we should get a better understanding of how executive power can
influence the risk of conflict onset and hopefully be able to
differentiate between anocratic regimes featuring similar levels of
state capacity and quality of elections.

## Executive Constraints

# Empirics {-}

## Research Design

To examine the relationship between civil conflict onset and executive
constraints, I construct a dataset of relevant variables and run a
bayesian multilevel logistic regression. In what follows, I first
describe the independent and dependent variables, including the
operationalizations and data sources, and then delve into the
regression model before finally presenting the results.

### Civil Conflict

The dependent variable, *civil conflict onset*, was taken from the
UCDP/PRIO Armed Conflict Dataset v19.1 (Cite) where intra-state
conflict is defined as a violent contest over a government or
territory between a state and non-state actor with a minimum threshold
of 25 battle deaths within a single country-year with a coverage from
1946 to 2018.  The data was merged based on country location for
conflicts labeled as internal or internationalized internal --- *i.e.*
where *type_of_conflict* was either 3 or 4. Multiple conflicts within
the same years were aggregated to form a binary variable coded as 1 to
denote the onset of at least one new civil conflict and 0 for all
other years, including years of ongoing conflict.^[Extrasystemic
conflicts, *i.e.* imperial or colonial conflicts, were excluded
largely due to lack of data. Ideally, such conflicts would be merged
based on the territory location of the conflict, however, for
colonies, for example, there is very little easily available economic
data and it would be undesirable to use the data for the colonizing
power due to the often wide discrepancies in development within
colonial empires. TODO: maybe a blurb on theory?]

The primary onset dates were determined according to UCDP/PRIO's
episodic definition where a new onset is coded for the same
incompatibility when there is at least a one year pause in the
fighting or drop below the battle death threshold. The primary
rationale is that civil conflicts often go through several distinct
phases of fighting which can make it difficult to accurately code the
end of a conflict. The episodic definition avoids potentially
conflating briefs lulls involving the same rebel groups and leaders
versus the re-emergence of previously long dormant
conflicts. Furthermore, in both scenarios, the theory linking
executive constraints to the incentives for political violence should
still hold (TODO: possible expand here).

That said, because of the potential sensitivity of results to changing
conflict definitions (cite Sambanis), all analyses were also run using
only unique onsets.^[Likewise, appendix ? presents results
using a higher threshold for battle deaths. The end results stay the
same TODO]

### Executive Constraints

(TODO: Probably cite Sklar and/or O'Donnell here for more theoretical
justification of operationalization, maybe even define institutions as
well)

To measure formal executive constraints I use data from the Varieties
of Democracy (V-Dem) dataset v9.0 (cite V-Dem). V-Dem is a
cross-national time series dataset providing multidimensional,
disaggregated data on democracy and political institutions for over
200 countries with an overall year coverage from 1789 until 2018. The
principal data is drawn from expert coded surveys involving a global
network of over 3000 coders, mostly academics as well as professionals
working in policy related fields (cite Methodology doc). Individual
ratings are aggregated into latent variable estimates using
hierarchical ordinal item response theory (IRT) models taking into
account rater reliability and varying rater thresholds (cite Pemstein
et al). ^[Thresholds relate to how the continuous latent trait for a
given variable is mapped on to the ordinal answer categories presented
on the survey. Due to differential item functioning, experts may
perceive the magnitude of difference between answer items
differently. For example, for the variable *v2elvotbuy*, election vote
buying, experts may vary in how they interpret the difference between
a 2 (Restricted Vote Buying) and a 3 (Almost No Vote Buying). These
two concepts, rater reliability and varying rater thresholds,
correspond respectively to the *discrimination* and *difficulty*
parameters in classical IRT models.] The result is over 220 lower
level indicators that are also combined through various aggregation
techniques to produce over 80 indices. Compared to datasets such as
Polity IV (cite), which has traditionally been used in most analyses
of civil conflict, V-Dem provides much more fine grained, nuanced
data, and most importantly, includes measures of uncertainty
propagated from the IRT models. This latter point is crucial since
ignoring measurement error, especially in models with noisy data and
small effect sizes as is often the case in conflict studies, can bias
estimates and lead to overconfident results (cite Treier & Jackman;
Kyle).

For the purposes of this paper, I re-estimate Lührmann *et al*'s
horizontal accountability index (*v2x_horacc*) in V-Dem with the
addition of several variables (cite). The *v2x_horacc* index aims to
capture the extent to which the executive is *de facto* constrained by
other government institutions --- more specifically, the legislature,
judiciary, constitutional norms, and independent investigative bodies
(cite pg 61) --- rather than the *de jure* powers of any one
institution *vis-à-vis* the executive. It is estimated as a latent
variable using a structural equation model thereby avoiding the
arbitrary weighting of linear additive indices.^[A theoretical
argument could be made that the oversight offered by each institution
is a necessary condition and that any index should therefore be
additive with equal weighting. However, this would be a normative
measure capturing the degree to which a country-year fulfills an ideal
concept, rather than measuring how constrained an executive actually
is in practice.]

For the legislature and the judiciary, the two principal government
branches most able to check the power of the executive, there needs to be
both a *capacity* and *willingness* for oversight. To better capture
the former dimension, I include two additional legislative
variables. Besides *v2lgqstexp*, legislature questions executive
branch officials, and *v2lginvstp*, legislature investigates in
practice, I add *v2lgfunds*, legislature controls its own resources,
and *v2lgoppart*, opposition parties able to exercise oversight. The
remaining manifest variables are unchanged. Namely, for the judicial
component: *v2juhcind*, high court independence; *v2juncind*, lower
court independence; *v2juhccomp*, the degree to which the government
complies with high court decisions; *v2jucomp*, the degree to which
the government complies with lower court decisions. Norms relating to
the status of the constitution are captured by *v2exrescon*, executive
respects constitution, and the final dimension, independent
investigative bodies, by *v2lgovost*, executive oversight by bodies
other than the legislature.

For the estimation of the latent index I largely follow the approach
taken by *v2x_horacc* with two notable exceptions. The first is that I
model the measurement error in the manifest variables using the
reported posterior standard deviations from V-Dem --- available as
_v2\*\_sd_. The second is that for computational reasons I allow the
judicial variables to impact the latent factor directly rather than
nested within a second latent factor (TODO: better justification). The
result is the following model specification.

For $m$ country-years indexed by $i \in I = \{1, \dotsc, n\}$ and $n$
manifest variables indexed by $p \in P = \{1, \dotsc, n\}$, let $Z \in
\mathbb{R}^{m \times n}$ be the matrix of manifest variables measured
with an observed degree of error, $Z^\ast \in \mathbb{R}^{m \times n}$
the matrix of latent true values, and $\theta$ the unidimensional
latent factor measuring executive constraints. Further, let $Q(Z_{\ast
p})$ be a property defined as true when $Z_{\ast p}$ denotes a
legislative variables and otherwise false.

I first model the measurement error in the manifest variables by
assuming that the observed values are drawn from a normal distribution
where the location parameters are the latent true values and the scale
the observed posterior standard deviations (cite McElreath).^[An even
more sensible approach would be to directly embed the individual IRT
models aggregating the manifest variables into the larger executive
constraints measurement model (cite Kyle). However, this prohibitively
increases the run time of the model.]
\[ Z_{ip} \sim \mathsf{Normal}(Z^\ast_{ip}, \tau_{ip}) \]

The latent values for the manifest variables are then modeled as
functions of the latent factor $\theta$. For the non-legislative
variables the mapping functions are identical to a standard factor
analytic approach.
\[
  Z^\ast_{ip} \sim \mathsf{Normal}(\gamma_p + \lambda_p \theta_i, \psi_p)
  \mbox{\quad if } p \in P \mid \neg Q(Z_{\ast p})
\]

where $\lambda$ is the variable specific factor loading, $\gamma$ the
population intercept, and $\psi$ the scale of the error terms.

In the case of the legislative variables there is the added issue of
missing observations for country-years lacking a
legislature.[^missing] For those observations, given the centrality of
the legislature as a potential constraint against the executive, the
latent factor should ideally be lower conditional on the other
manifest variables. However, since the manifest variables are
themselves latent variables, minimal values cannot be imputed without
a nonlinear transformation to a bounded scale. I thus follow
*v2x_horacc* in mediating the effect of the latent factor on the
legislative variables by the presence of the legislature using a
dichotomized version of *v2lgbicam*, how many chambers does the
legislature contain, denoted by the Bernoulli distributed variable,
$w$, where country-years with a sitting legislature are given a $1$
and country-years where the legislature is missing or has been
suspended, $0$. The probability of a legislature being present,
$\pi_i$, is then a linear function of the latent factor, $\theta$,
transformed by the inverse logistic such that $\pi_i \in [0,
1]$.^[Technically, the inverse logistic transforms values to the open
interval, $(0, 1)$, but due to overflow/underflow in floating point
calculations by computers the endpoints can be observed.] The
legislative manifest variables are in turn treated as functions of the
linear predictor for $w$ (cite pg 63). The practical effect is that
where there is data for the legislative manifest variables, the latent
factor is informed by the measured institutional quality of the
legislature. Where data is missing due to the suspension or lack of a
legislature, the latent factor is down weighted by the linear function
predicting $w$, conditional on the observed values of the remaining
manifest variables.
\begin{align*}
  w_i & \sim \mathsf{Bernoulli}(\pi_i) \\
  \pi_i & = \mathsf{logit}^{-1}(\xi_i) \\
  \xi_i & = \eta + \delta \theta_i \\
  Z^\ast_{ip} & \sim \mathsf{Normal}(\gamma_p + \lambda_p \xi_i, \psi_p)
    \mbox{\quad if } p \in P \mid Q(Z_{\ast p}) \land w_i = 1
\end{align*}

[^missing]: Missingness due to the lack of of a particular
institution, for example the suspension of high court, is an issue for
the other manifest variables as well. However, only the existence of
the legislature is coded by V-Dem as a separate variable. Furthermore,
practically speaking, none of the other manifest variables exhibit the
same systematic pattern of missingness in the dataset as the
legislative variables. Whereas there are 1608 country-years lacking a
legislature ($13.4\%$ of all post-1945 country-years), only 115 are
missing data for at least one of the judicial variables ($1\%$) and
only 58 are missing for *v2exrescon* ($0.5\%$). The variable
*v2lgotovst*, executive oversight by bodies other than the
legislature, is an outlier with 1619 country-years missing ($13.5\%$)
since there is an error in V-Dem's v9.0 dataset whereby it has been
cleaned according to when the legislature is missing. Thus, I impute
*v2lgotovst* within the larger model where the legislature is missing
by drawing missing values from a normal distribution with a mean and
standard deviation estimated from the observed *v2lgotovst* data. No
other country-years or variables are imputed. The rationale is that,
disregarding data processing issues, missingness in V-Dem can arise
either from a lack of expert coders or lack of information. The latter
scenario will most likely be correlated with country-years where there
is a crisis or significant political instability, such as coups or
civil war, where extreme values are more probable. Since it cannot be
assumed that the values are completely missing at random, the data
generative process for missingness cannot be modeled without
potentially introducing bias into the estimates.

Finally, for the both the legislative and non-legislative mapping
functions I assume local item independence such that the elements of
$\Psi$ are uncorrelated (cite Lord & Novick). Furthermore, for the
purposes of identifying the likelihood the scale of the latent factor,
$\theta$, is fixed to mean $0$ and standard deviation $1$ while the
slopes, $\Lambda$ and $\delta$, are constrained to be positive (cite
Jackman). The positive constraints are drawn from theory since all of
the manifest variables exhibit the same scale direction and should be
positively correlated with the latent factor. Similarly, the presence
of the legislature should be positively correlated with higher values
of the latent factor, *i.e.* greater executive constraints.

### Controls

In addition to the data on conflict onset and executive constraints, I
also include a number of standard controls drawn largely from previous
literature, the most important of which is state capacity. According
to the theory outlined earlier, the impact of executive constraints on
decreasing the probability of conflict onset should be conditional on
the relative capacity of the state. The lack of executive constraints
most likely increases the risk of civil conflict when state capacity
is low and the central government is willing to use repression, but
lacks the ability to fully suppress the opposition. To measure state
capacity I follow Fearon & Laitin in using GDP per capita as a proxy
for the state's administrative and military capacity by roughly
capturing the revenue potential available to the state (cite). As a
data source, I use the natural log of the *RGDP^e* variable, constant
price real GDP, from Penn World Tables v9.1 (1950-2017) to allow for
both cross-country and cross-temporal comparisons of GDP per capita
(cite)^[TODO: Expand on why Penn World Tables] and include an
interaction term between GDP per capita and the latent measure for
executive constraints to capture the conditional relationship.



- state capacity (GDP per capita)

- gdp growth

- population density

- mean elevation

- relevant group counts

- neighbouring conflict

- ongoing conflict

- peace years

- varying intercepts

What's not included (political instability) and why

### Regression Model

I model civil conflict onset as a bayesian multilevel regression where
onset is a function of a set of predictors, including executive
constraints and an array of control variables, together with several
grouping factors included as partially pooled varying intercepts. The
units of analysis are all independent country-years from 1946 until
2016, as determined by data availability, with *n* countries indexed
by $i = 1 \dotsc n$ and *m* years indexed by $t = 1 \dotsc m$. The
dependent variable, conflict onset, is a Bernoulli distributed binary
variable where 1 denotes the onset of at least one new civil conflict
in a given country-year. The chosen link function transforming the
linear regression output into the probability of conflict onset,
$\nu_{it}$, is the inverse logistic such that $\nu_{it} \in [0, 1]$.

The full regression specification is then as follows:
\begin{align*}
  y_{it} & \sim \mathsf{Bernoulli}(\nu_{it}) \\
  \nu_{it} & = \mathsf{logit^{-1}}(\alpha + X_{it-1} \beta + \sum_{k=1}^K \zeta^k_{j_{it}})
\end{align*}

where **X** is a single row matrix of predictors, $\beta$ a vector of
regression slopes, and $\delta^k$ the set of varying intercepts for
*K* grouping factors indexed by $j_k$. To account for time
endogeneity, all predictors are lagged by one year. They are also
standardized by subtracting the mean and dividing by the standard
deviation for each variable to aid in computation as well as
comparison of the final $\beta$ estimates (cite Gelman maybe?).

I place weakly informative priors on all parameters to allow the
posterior estimates to be largely determined by the observed data
while avoiding the more pathological values and hard constraints that
can arise from uninformative priors (cite Stan). For the slopes and
population intercept, given that the input data has been standardized,
this translates into

\[ \alpha \sim \mathsf{Normal}(0, 5) \]
\[ \beta \sim \mathsf{Normal}(0, 2.5) \]

Finally, for the varying intercepts modeling group-specific
variability, I use hierarchical priors that provide a regularizing
effect against overfitting (cite Gelman and Hill - Multilevel
Regression).

\[ \delta^k \sim \mathsf{Normal}(0, \sigma^k) \]
\[ \sigma^k \sim \mathsf{HalfCauchy}(0, 1) \]

## Analysis

Measurement model:
\begin{align*}
  \theta & \sim \mathsf{Normal}(0, 1) \\
  \delta, \lambda & \sim \mathsf{LogNormal}(0, 1) \\
  \eta, \gamma & \sim \mathsf{Normal}(0, 5) \\
  \psi & \sim \mathsf{Weibull(2, 1)}
\end{align*}

Regression parameters:
\begin{align*}
  \beta & \sim \mathsf{Normal}(0, 2.5) \\
  \alpha & \sim \mathsf{Normal}(0, 5) \\
  \zeta^k & \sim \mathsf{Normal}(0, \sigma^k) \\
  \sigma & \sim \mathsf{HalfCauchy}(0, 1)
\end{align*}


```{r}
beta <- readRDS("posteriors/summary/beta.rds")
plot_pars(beta)
```

```{r}
input.df <- readRDS("posteriors/input_data.rds")
p_hat <- readRDS("posteriors/summary/predicted_probs.rds")

plot_dens(p_hat, groups = input.df$lepisode_onset)
```


```{r}
p <- apply(p_hat, 2, median)
x <- evalmod(scores = p, labels = input.df$lepisode_onset)
plot(x, curvetype = c("ROC", "PRC"))
```

```{r eval = F}
# Interaction plot
intercept <- readRDS("posteriors/summary/intercept.rds")
theta <- readRDS("posteriors/summary/theta.rds") %>% apply(2, median) %>% sort

state <- quantile(input.df$cgdppc, probs = c(0.2, .5, 0.8))

f <- function(x, y) {
    with(input.df, intercept + beta[, 1] * x + beta[, 2] * y +
                   beta[, 3] * x * y + beta[, 4] * mean(peace_yrs) +
                   beta[, 5] * mean(gdpgro) + beta[, 6] * mean(pop_density)  +
                   beta[, 7] * mean(meanelev) + beta[, 8] * mean(rlvt_groups_count) +
                   beta[, 9] * 1 + beta[, 10] * 0)
}

par(mfrow = c(1, 3))
for (y in state) {
    print(y)
    p <- boot::inv.logit(sapply(theta, f, y)) * 100

    m <- apply(p, 2, median)
    pi <- apply(p, 2, quantile, probs = c(0.2, 0.8)) %>% t

    plot(theta, m, type = "l", ylim = c(min(pi[, 1]), max(pi[, 2])),
         xlab = "Executive Constraints", ylab = "Predicted Probability")
    lines(theta, pi[, 1], lty = 2)
    lines(theta, pi[, 2], lty = 2)
}


```

# Conclusion {-}

# Appendices {-}

## Appendix A - Stan Code {#appendix-a}

```{stan, code = readLines("stan/model.stan"), eval = FALSE, output.var = "x", echo = T}
```

\newpage

## Appendix B - Prior Predictive Check {#appendix-b}


# References {-}

\singlespacing
